Analysing Constraint Grammars with a SAT-solver

### 2016-03-02 paper/lrec/eval_futurework.tex
little cleanup

\paragraph{Accuracy} 
The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not check for false negatives in any of the grammars. However, the test reported a high number of errors even in the smallest grammars, and some of them are intuitively not that bad; we

@@

\begin{itemize}
\item[] 
\begin{verbatim}REMOVE Adv IF (1 Noun) ;
REMOVE Adv IF (-1 Det) (0 Adj) (1 Noun) ;
\end{verbatim}
\end{itemize}

@@

As per VISL CG-3, the condition \texttt{(0 "zijn")} does not require that the input would have determiner and ``zijn'' in the same reading: just that there is any determiner and any ``zijn'' in the cohort.

\todo{Find a better example of the imprecise 0 condition --- the ambiguity class thing would eliminate this one!}

Can we catch this imprecise formulation with our tool? The example word constructed by the SAT-solver may as likely be one of the following:

@@

\item[b.] \begin{verbatim}
"w2"
    w2<det><pos><mfn><pl><"zijn">
    w2<anything>
\end{verbatim}
\end{itemize}

Incidentally, the first example of a rule conflict had a similar requirement, but used correctly: the intended semantics was indeed that the adjective and the adverb are in different readings. However, in this case, the correct way to express the target is \texttt{SELECT DetPos + "zijn"}.

In the latter case, the grammar writer would hopefully notice the imprecise definition and change their definition. However, this is more of a happy side effect than intended feature---the grammar writer cannot count on our tool to notice all similar cases.
This kind of confusion is found in other rules as well. 
To catch them more systematically, we could add a feature that alerts in all cases where a condition with 0 is used, and automatically construct a version of the rule that moves whatever tags in the 0-condition into the target, then asks the user which one was meant.

@@

In addition, there was a number of definitions that were never
used. The program doesn't currently point them out to the user, but
that is a trivial addition; just like rule-internal conflicts, it does
not require any semantic analysis. 
Another error, which was not pointed by the program, but discovered when we tried to extract the readings from the grammar and use them

@@

\begin{itemize}
\item[] 
\begin{verbatim}
REMOVE Podar  IF (0 Podar) (1 Adv) (2C Inf OR Enc) ;
REMOVE Sentar IF (0 Sentar) (1C Inf OR Enc) ;
\end{verbatim}
\end{itemize}

@@

\paragraph{Finnish} 
\todo{Rerun the tests and rewrite this whole section!}

>>>>

The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not check for false negatives in any of the grammars. However, the test reported a high number of errors even in the smallest grammars, and some of them are intuitively not that bad; we

@@

\begin{itemize}
\item[] 
\begin{verbatim}REMOVE Adv IF (1 N) ;
REMOVE Adv IF (-1 Det) (0 Adj) (1 N) ;
\end{verbatim}
\end{itemize}

@@

As per VISL CG-3, the condition \texttt{(0 "zijn")} does not require
that the input would have determiner and ``zijn'' in the same reading:
just that there is a reading with any determiner, and a reading with any ``zijn'' in the cohort. 
Can we catch this imprecise formulation with our tool? The example word constructed by the SAT-solver may as likely be one of the following:

@@

\item[b.] \begin{verbatim}
"w2"
    w2<det><pos><mfn><pl><"zijn">
\end{verbatim}
\end{itemize}

Incidentally, the first example of a rule conflict had a similar
requirement, but used correctly: the intended semantics was indeed
that the adjective and the adverb are in different readings. However,
in this case, the precise way to express the target is \texttt{SELECT DetPos + "zijn"}.

% In the latter case, the grammar writer would hopefully notice the imprecise definition and change their definition. However, this is more of a happy side effect than intended feature---the grammar writer cannot count on our tool to notice all similar cases.

@@

In addition, there was a number of definitions that were never
used. The program doesn't currently point them out to the user, but
that is a trivial addition; just like rule-internal conflicts, it does
not require any semantic analysis. 
We also discovered problems in some set definitions, when we tried to
use the set definitions from the grammar directly as our readings.
For instance, the following definition requires the word to be all of
the listed parts of speech at the same time:
\begin{itemize}
\item[] 
\texttt{SET NP\_Member = N + A + Det + PreAdv + Adv + Pron ;}
\end{itemize}

This particular definition was never used in the rules, so it would not
have caused a conflict. We noticed it by accident, when the SAT-solver
offered it in an example sequence that, meant for another rule.

@@

\begin{itemize}
\item[] 
\texttt{REMOVE Podar  IF (0 Podar) (1 Adv) (2C Inf OR Enc) ;} \\
\texttt{REMOVE Sentar IF (0 Sentar) (1C Inf OR Enc) ;}
\end{itemize}

@@

We found the same kind of formulation in many other rules and grammars.
To catch them more systematically, we could add a feature that alerts in all cases where a condition with 0 is used, and automatically construct a version of the rule that moves whatever tags in the 0-condition into the target, then asks the user which one was meant.

\paragraph{Finnish}

\todo{Rerun the tests and rewrite this whole section!}

### 2016-03-02 paper/lrec/implementation.tex
add small table about ambiguity classes
little cleanup

We solve the problem by creating \emph{ambiguity classes}: groups of analyses that can be ambiguous with each other. 
We represent the expanded morphological lexicon as a matrix, as seen in figure~\todo{make a nice picture with rows and columns}: word forms on the rows and analyses on the columns. Each distinct row forms an ambiguity class. For example, the ambiguity class \verb![1257,496,16]! %TODO better description
contains masculine plural adjectives and masculine plural past participles.
Then we form SAT clauses that allow or prohibit certain combinations. These clauses will interact with the constraints created from the rules, and the end result will be closer to real-life sentences.

>>>>

\begin{center}
\begin{tabular}{c|c|c|c|c}

            & n nt sg  & n f pl  & vblex sep inf & det pos mfn  \\ \hline
uitgaven    & 0        & 1       & 1             & 0    \\ 
toespraken  & 0        & 1       & 1             & 0    \\ 
haar        & 1        & 0       & 0             & 1    \\

\end{tabular}
\end{center}

We solve the problem by creating \emph{ambiguity classes}: groups of analyses that can be ambiguous with each other. 
We represent the expanded morphological lexicon as a matrix, as seen
above: word forms on the rows and analyses on the columns. Each
distinct row forms an ambiguity class. For example, one class may
contain words that are ambiguous between plural feminine nouns and
separable verb infinitives; another contains masculine plural adjectives 
and masculine plural past participles.
Then we form SAT clauses that allow or prohibit certain combinations. These clauses will interact with the constraints created from the rules, and the end result will be closer to real-life sentences.

### 2016-03-02 paper/lrec/problem_desc.tex
minor tweaking so that all bullet points are on the same page
little cleanup

\begin{itemize}
\item Are all the rules distinct? (e.g. \texttt{Para} and \texttt{De} may be included in \texttt{Prep})
\item Can two or more rules be merged? (e.g. \texttt{SELECT Inf IF
    (-1C Prep OR Vai OR Vbmod)})
\item Does my neat rewrite of a messy rule preserve the meaning?
\item What is the best order for the rules?
\item Generate a sequence that triggers rules $R$ but not rules $R'$ 
\end{itemize}

@@

The paper is structured as follows. Section~\ref{sec:prev} relates our
work to the previous research. 
Section~\ref{sec:implementation} discusses the implementation more in depth, and
Section~\ref{sec:eval} presents preliminary results. 
Section~\ref{sec:conclusion} discusses future work and concludes the paper.

>>>>

\begin{itemize}
\item Are all the rules distinct? (e.g. \texttt{Para} and \texttt{De} may be included in \texttt{Prep})
\item Can two or more rules be merged? (e.g. \texttt{SELECT Inf IF
    (-1C Prep OR Vai OR Vbmod)})
\item Can I rewrite a messy rule in a neater way without changing the meaning?
\item What is the best order for the rules?
\item Generate a sequence that triggers $R$ but not $R'$ 
\end{itemize}

@@

The paper is structured as follows. Section~\ref{sec:prev} relates our
work to the previous research. 
Section~\ref{sec:implementation} discusses the implementation, and
Section~\ref{sec:eval} presents preliminary results. 
Section~\ref{sec:conclusion} discusses future work and concludes the paper.
